---
title: "var-2-notebook"
format: html
editor: visual
---

## Loading packages

```{r}
library(ggplot2)
library(tidyverse)

library(cmdstanr)
library(posterior)
library(bayesplot)

# MASS::mvrnorm() is called qualified: attaching MASS would mask dplyr::select().
```

```{r}
# On the DEAC cluster, point cmdstanr at the shared install:
# set_cmdstan_path("/deac/sta/classes/sta720/software/cmdstan/2.37.0")
```

### Helpers for reading matrix-valued parameters

`fit$summary()` returns one row per scalar element (`A_1[1,2]`, `A[2,1,2]`, ...). These pull the posterior means back into the matrix shape of the simulation parameters, indexing explicitly so the result does not depend on the order cmdstan writes the elements in.

```{r}
post_matrix <- function(fit, name, K) {
  s   <- fit$summary(name)
  idx <- str_match(s$variable, "\\[(\\d+),(\\d+)\\]")
  out <- matrix(NA_real_, nrow = K, ncol = K)
  out[cbind(as.integer(idx[, 2]), as.integer(idx[, 3]))] <- s$mean
  out
}

post_array <- function(fit, name, P, K) {
  s   <- fit$summary(name)
  idx <- str_match(s$variable, "\\[(\\d+),(\\d+),(\\d+)\\]")
  out <- array(NA_real_, dim = c(P, K, K))
  out[cbind(as.integer(idx[, 2]), as.integer(idx[, 3]), as.integer(idx[, 4]))] <- s$mean
  out
}
```

### Simulating VAR(2) data

```{r}
sim_var_model <- function(c_vec, A_1, A_2, Sigma, T_obs, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  stopifnot(T_obs >= 3)

  K <- length(c_vec)
  Y <- matrix(0, nrow = T_obs, ncol = K)

  epsilon <- MASS::mvrnorm(n = T_obs, mu = rep(0, K), Sigma = Sigma)

  # Unconditional mean of the stable VAR(2)
  mu <- solve(diag(K) - A_1 - A_2, c_vec)
  Y[1, ] <- MASS::mvrnorm(1, mu, Sigma)
  Y[2, ] <- MASS::mvrnorm(1, mu, Sigma)

  for (t in 3:T_obs) {
    Y[t, ] <- c_vec + A_1 %*% Y[t - 1, ] + A_2 %*% Y[t - 2, ] + epsilon[t, ]
  }
  Y
}
```

Checking that the VAR(2) selected is stable, by confirming that every eigenvalue of the companion matrix lies inside the unit circle.

```{r}
T_obs <- 200
c_vec <- c(2.5, 1.5)
A_1 <- matrix(c(0.45, -0.15,
                -0.35, 0.30), nrow = 2)
A_2 <- matrix(c(0.05, -0.05,
                -0.05, 0.03), nrow = 2)
Sigma <- matrix(c(1.0, 0.1,
                  0.1, 1.0), nrow = 2)

K <- length(c_vec)

companion <- rbind(
  cbind(A_1, A_2),
  cbind(diag(K), matrix(0, nrow = K, ncol = K))
)
Mod(eigen(companion)$values)
```

Simulating, plotting, and saving the synthetic series. The CSV is written without a header to match how `var-2-conjugate-litterman.qmd` reads it.

```{r}
Y <- sim_var_model(c_vec, A_1, A_2, Sigma, T_obs, seed = 1)

dir.create("simdata", showWarnings = FALSE)
write.table(Y, "simdata/Y_sim.csv", sep = ",",
            row.names = FALSE, col.names = FALSE)

sim_long <- as.data.frame(Y) %>%
  setNames(c("y1", "y2")) %>%
  mutate(t = row_number()) %>%
  pivot_longer(c(y1, y2), names_to = "series", values_to = "value")

ggplot(sim_long, aes(t, value, color = series)) +
  geom_line() +
  labs(x = "t", y = NULL, color = NULL,
       title = "Simulated VAR(2) series") +
  theme_minimal()
```

### Fitting the VAR(2) model to simulated data

```{r}
stan_data <- list(
  T = T_obs,
  K = K,
  Y = Y
)

mod <- cmdstan_model(
  "models/var-2.stan",
  stanc_options = list("O1", "warn-pedantic" = TRUE)
)

fit <- mod$sample(
  data = stan_data,
  refresh = 0,
  show_messages = FALSE,
  show_exceptions = FALSE
)
```

### Posterior analysis of model

```{r}
fit$diagnostic_summary()
fit$summary(c("c", "A_1", "A_2", "sigma"))
```

```{r}
cat("True c:\n");        print(c_vec)
cat("Estimated c:\n");   print(fit$summary("c")$mean)

cat("True A_1:\n");      print(A_1)
cat("Estimated A_1:\n"); print(post_matrix(fit, "A_1", K))

cat("True A_2:\n");      print(A_2)
cat("Estimated A_2:\n"); print(post_matrix(fit, "A_2", K))
```

```{r}
dir.create("fits", showWarnings = FALSE)
fit$save_output_files(dir = "fits", basename = "var2",
                      timestamp = FALSE, random = FALSE)
```

### Fitting the hierarchical Minnesota model to simulated data

```{r}
P <- 2

stan_data_hier <- list(
  T = T_obs,
  K = K,
  P = P,
  Y = Y
)

mod_hier <- cmdstan_model(
  "models/var-2-hierarchical.stan",
  stanc_options = list("O1", "warn-pedantic" = TRUE)
)

fit_hier <- mod_hier$sample(
  data = stan_data_hier,
  refresh = 0,
  show_messages = FALSE,
  show_exceptions = FALSE
)
```

### Posterior analysis of model

```{r}
fit_hier$diagnostic_summary()
fit_hier$summary(c("c", "A", "sigma", "lambda", "theta"))
```

```{r}
A_hat <- post_array(fit_hier, "A", P, K)

cat("True c:\n");        print(c_vec)
cat("Estimated c:\n");   print(fit_hier$summary("c")$mean)

cat("True A_1:\n");      print(A_1)
cat("Estimated A_1:\n"); print(A_hat[1, , ])

cat("True A_2:\n");      print(A_2)
cat("Estimated A_2:\n"); print(A_hat[2, , ])
```

```{r}
dir.create("fits", showWarnings = FALSE)
fit_hier$save_output_files(dir = "fits", basename = "var2-hierarchical",
                           timestamp = FALSE, random = FALSE)
```
