---
title: "VAR(2) with Conjugate Priors"
format: html
editor: visual
---

## Loading packages

```{r}
library(ggplot2)
library(tidyverse)
library(jsonlite)
```

### Loading simulated data

We're going to run some Bayesian VAR(2) models in R (conjugate priors and Minnesota with fixed innovation matrix).

`simdata/Y_sim.csv` is written by `var-2-notebook.qmd`; run that notebook first.

```{r}
Y <- as.matrix(read.csv("simdata/Y_sim.csv", header = FALSE, col.names = c("y1", "y2")))

T <- nrow(Y) # Number of time periods
K <- ncol(Y) # Number of variables
P <- 2

Y_eff <- Y[(P+1):T, ]                    
X <- cbind(1,
           Y[P:(T-1), ],                 
           Y[1:(T-2), ])                 
                                      
```

### Conjugate Minnesota posterior for fixed hyperparameters

This block sets the "known" sigma's from the data by taking the standard deviation of the residuals for each of the 2 univariate AR processes:

```{r}
lambda <- 0.2  
theta  <- 0.2  
alpha    <- 1   
delta  <- rep(1, K)
sigma <- numeric(K)

for (k in 1:K) {
  sigma[k] <- sqrt(arima(Y[, k], order = c(P, 0, 0))$sigma2)
}

M <- K * P + 1
B_post     <- matrix(0, nrow = M, ncol = K)
V_post_lst <- vector("list", K)

for (j in 1:K) {
    # Step 2: build gamma_bar, tau
    gamma_bar <- numeric(M)
    tau       <- numeric(M)
    tau[1]    <- 1e6

    for (l in 1:P) {
      for (r in 1:K) {
        pos <- 1 + (l - 1) * K + r
        if (r == j) {
          gamma_bar[pos] <- if (l == 1) delta[j] else 0
          tau[pos]       <- lambda / l^alpha
        } else {
          gamma_bar[pos] <- 0
          tau[pos]       <- (lambda * theta * sigma[j]) / (l^alpha * sigma[r])
        }
      }
    }
  
    # Step 3: build R_j, r_j
    R_j <- diag(sigma[j] / tau)        # M x M diagonal
    r_j <- gamma_bar * (sigma[j] / tau) # length M
    
    # Step 4: compute posterior moments
    XtX_post   <- crossprod(X) + crossprod(R_j)
    gamma_post <- solve(XtX_post, crossprod(X, Y_eff[, j]) + crossprod(R_j, r_j))
    V_post     <- sigma[j]^2 * solve(XtX_post)
    
    B_post[, j]     <- gamma_post
    V_post_lst[[j]] <- V_post
}

print(B_post)
```

### Wrapping the posterior in a function

```{r}
conj_litter <- function(Y, P, lambda, theta, alpha = 1) {
  
  T <- nrow(Y) # Number of time periods
  K <- ncol(Y) # Number of variables
  P <- P # Number of lags

  # NOTE: X is written out for two lags only, so this is P = 2 despite the
  # argument; generalizing it means building the lag block in a loop over P.
  Y_eff <- Y[(P+1):T, ]                    
  X <- cbind(1,
            Y[P:(T-1), ],                 
            Y[1:(T-2), ])  
  
  lambda <- lambda  
  theta  <- theta  
  alpha  <- alpha   
  delta  <- rep(1, K)
  sigma <- numeric(K)
  
  for (k in 1:K) {
    sigma[k] <- sqrt(arima(Y[, k], order = c(P, 0, 0))$sigma2)
  }

  M <- K * P + 1
  B_post     <- matrix(0, nrow = M, ncol = K)
  V_post_lst <- vector("list", K)

  for (j in 1:K) {
      # Step 2: build gamma_bar, tau
      gamma_bar <- numeric(M)
      tau       <- numeric(M)
      tau[1]    <- 1e6

      for (l in 1:P) {
        for (r in 1:K) {
          pos <- 1 + (l - 1) * K + r
          if (r == j) {
            gamma_bar[pos] <- if (l == 1) delta[j] else 0
            tau[pos]       <- lambda / l^alpha
          } else {
            gamma_bar[pos] <- 0
            tau[pos]       <- (lambda * theta * sigma[j]) / (l^alpha * sigma[r])
          }
        }
      }
    
      # Step 3: build R_j, r_j
      R_j <- diag(sigma[j] / tau)        # M x M diagonal
      r_j <- gamma_bar * (sigma[j] / tau) # length M
      
      # Step 4: compute posterior moments
      XtX_post   <- crossprod(X) + crossprod(R_j)
      gamma_post <- solve(XtX_post, crossprod(X, Y_eff[, j]) + crossprod(R_j, r_j))
      V_post     <- sigma[j]^2 * solve(XtX_post)
      
      B_post[, j]     <- gamma_post
      V_post_lst[[j]] <- V_post
  }
  return(t(B_post))
}

```

### Sweeping over the tightness hyperparameter

```{r}
lambda_check <- c(0.1, 0.2, 0.3, 0.4, 0.5)
B_post_list <- vector(mode = "list", length = length(lambda_check))

for (i in 1:length(lambda_check)) {
   coef <- conj_litter(Y, P = 2, lambda = lambda_check[i], theta = 0.2, alpha = 1)

   B_post_list[[i]] <- coef
}

B_post_list[1]
```

### Posterior coefficients vs. tightness

```{r}
dir.create("figures", showWarnings = FALSE)

coef_names <- c("const", "y1.l1", "y2.l1", "y1.l2", "y2.l2")
eq_names   <- c("y1", "y2")

coef_df <- map2(B_post_list, lambda_check, function(B, lam) {
  colnames(B) <- coef_names
  as_tibble(B) %>%
    mutate(equation = eq_names, lambda = lam)
}) |>
  list_rbind() %>%
  pivot_longer(all_of(coef_names), names_to = "param", values_to = "value")

p_conj <- coef_df %>%
  filter(param != "const") |>  # const has a near-flat prior (τ=1e6); drop so AR scale is readable
  ggplot(aes(lambda, value, color = param)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ equation, labeller = label_both) +
  labs(x = expression(lambda), y = "Posterior mean coefficient",
       color = "Coefficient",
       title = "Minnesota tightness vs. posterior coefficients") +
  coord_cartesian(xlim = c(0, 0.85), ylim = c(-0.5, 1.1)) +
  theme_minimal()

p_conj
ggsave("figures/coef_conjugate.png", p_conj, width = 9, height = 4, dpi = 150)
```

### Comparing against the hierarchical Stan sweep

`results/lambda_rate` holds one subdirectory per swept gamma-prior rate, each
with `metadata.json` and `summary.csv` from the cluster run.

```{r}
sweep_dir <- "results/lambda_rate"

hier_coef <- list.dirs(sweep_dir, recursive = FALSE) |>
  map(function(d) {
    meta <- fromJSON(file.path(d, "metadata.json"))
    summ <- read_csv(file.path(d, "summary.csv"), show_col_types = FALSE)
    names(summ)[1] <- "param"
    summ |>
      filter(str_detect(param, "^A\\[")) |>
      transmute(param, value = Mean,
                prior_mean = meta$lambda_prior_mean,
                post_lambda = meta$lambda_post_mean)
  }) |>
  list_rbind() |>
  extract(param, into = c("lag", "eq", "var"),
          regex = "A\\[(\\d+),(\\d+),(\\d+)\\]", convert = TRUE) |>
  mutate(equation = paste0("y", eq),
         param    = paste0("y", var, ".l", lag))

p_hier <- hier_coef |>
  ggplot(aes(prior_mean, value, color = param)) +
  geom_line() + geom_point() +
  facet_wrap(~ equation, labeller = label_both) +
  labs(x = expression(lambda ~ "prior mean (shape/rate)"), y = "Posterior mean coefficient",
       color = "Coefficient", title = "Hierarchical Stan") +
  coord_cartesian(xlim = c(0, 0.85), ylim = c(-0.5, 1.1)) +
  theme_minimal()

p_hier
ggsave("figures/coef_hierarchical.png", p_hier, width = 9, height = 4, dpi = 150)
```